\name: A flexible and data-driven $K$-process model for nucleosynthesis

### 2023-06-04 Paper/main.tex
removed half of the Hoggs

% typesetting stuff
\shorttitle{a data-driven model for nucleosynthesis}
\shortauthors{Griffith \& Hogg}
\addtolength{\topmargin}{-0.4in} % trust in Hogg
\addtolength{\textheight}{0.7in} % trust in Hogg
\setlength{\parindent}{1.6em}
\renewcommand{\twocolumngrid}{} % once again
\renewcommand{\paragraph}[1]{\bigskip\par\noindent{\textbf{#1}}~---}
\sloppy\sloppypar\raggedbottom\frenchspacing % Trust Hogg

@@

\begin{abstract}\noindent % trust Hogg
Stellar surface abundances look like they are produced by two dominant processes, one prompt and one delayed.
We analyze the abundances of 14 elements in [X/Mg] vs. [Mg/H] space for 48,659 stars from APOGEE DR17 with a flexible, data-driven $K$-process model---dubbed \name.
In the fiducial model, where $K=2$, each elemental abundance of each star is described as the vector sum of a prompt (CCSN-like) and delayed (SNIa-like) process.
While prior work derives process vector components from median observed high-Ia and low-Ia abundance trends, we relax these assumptions and simultaneously fit all model parameters to the data, with only minimal constraints to keep the processes interpretable.
We find that our flexible method is able to recover the observed abundance bimodality and that this bimodality is visible to higher [Mg/H] in process amplitude space than in abundance space.
We compare fit parameters and statistics from this new method to those from prior work, finding that the two methods produce similar results, but that \name{} better predicts stellar abundances, especially for C+N and Mn.
We test the validity of the standard two-process model assumptions, such as that Fe is generated by a roughly half-half combination of prompt and delayed sources.
We find that the assumptions of prior works are generally confirmed, but that the inferred fractional contributions from each process are strongly dependent upon Fe assumptions.
We also explore third and fourth processes, calibrated to Ce and Mn, respectively. The additional processes improve the model's ability to predict the stellar abundances, demonstrating the model's ability to extend beyond $K=2$. \name{} enables future research to study the disk's two-dimensionallity, investigate age-abundance label correlations, fit non-bimodal abundance distributions with a multi-component nucleosynthetic model, and much more. \ejg{Add something about code availability.} 
\end{abstract}

@@

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data. In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment. The parameters output by our model can be used as de-noised abundance label to sharpen relationships between abundances and stellar parameters.  \ejg{Add more about high-quality metallicity labels? Hogg to edit this paragraph.}

@@

\paragraph{1. $K$ processes}
All elements on the periodic table are produced by a combination of nucleosynthetic processes such as CCSN, SNIa, AGB stars, and merging neutron stars \citep{johnsonja2020}. The majority of $\alpha$, light odd-Z, and Fe-peak elements (the elements observed by APOGEE) are dominantly produced by $K=2$ sources, with one being a prompt process or mix of prompt processes, and one being a delayed process or a mix of delayed processes. This is substantiated by theoretical yields \citep[e.g.,][]{anderson2019, rybizki2017} and past successful data-driven models (e.g., G22, W22, \citealp{ting2022}). In this paper we therefore assume that $K \geq 2$, though \name{} could be implemented with $K=1$. \hogg{This assumption says too much; we could dial this back.}

\paragraph{2. Linearity}
At every metallicity, the (X/H) abundances of a star can be expressed as a linear combination of $K$ processes.
These processes themselves will depend on metallicity, but a linear sum is sufficient to explain all element abundances at any overall metallicity.
Because yields depend on stellar masses and progenitor structure, and because different stars can get to their metallicities by different histories, this assumption must be at least slightly wrong in detail.

@@

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
The APOGEE errors are likely slightly underestimated for some element measurements, so we add softening parameter $Q$ to inflate the errors (see Equation~\ref{eq:inflate_ivar}). \hogg{Make this more positive about apogee}

@@

While \name{} is flexible, it still requires assumptions to break model degeneracies (see Section~\ref{sec:model}). Specifically, it requires an element with ``known'' $\qcc$ and an element with known $\qIa$ to constrain each process. \ejg{Hogg edit this paragraph.} For the prompt process, we ground our assumption in the nucleosynthetic theory that Mg is a pure CCSN element \citep[e.g.,][]{andrews2017}. Unfortunately there is no comparable pure SNIa element, nor is there an element for which we know the relative CCSN/SNIa ratio. In order to regularize the delayed process, we choose to fix the Fe process vector components, making unsubstantiated assumptions about Fe enrichment.

@@

We fit \name{} with $K=2$ to 15 abundances labels for 48,659 RGB stars in APOGEE DR17, selecting a population that minimizes statistical and systematic errors while spanning an [Mg/H] of -0.8 to 0.5. In the $K=2$ model, the first process, fixed to Mg, represents prompt CCSN-like enrichment, and the second process, fixed to Fe, represents delayed SNIa-like enrichment---but other nucleosynthetic sources with similar time delays may be mixed into each. Overall, we find that $K=2$ is a good fit to the data and that the model successfully recovers the global abundance patterns in the Milky Way. While \name{} does not rely on $\femg$ vs. $\mgh$ bimodality or median abundance trends, it is able to recover the observed bimodal abundance distribution. Further, the fit parameters $\Acc$ and $\AIa$ act as ``denoised'' abundance labels, and reveal a clearer signature of bimodality at high-metallicity in $\AIa/\Acc$ vs. $\mgh$ space than in $\femg$ vs. $\mgh$. This suggests that the $KPM$ fit parameters and predicted abundances could be used as a higher signal-to-noise tracer of nucleosynthesis, as they are condensing information from 15 elements into two variables.

@@

Finally, we expand \name{} from $K=2$ to $K=4$, regularizing the third and fourth processes to Ce and Mn. KPM builds off of the original model, such that the $K=4$ model starts at the $K=2$ solution and then finds the best-fit parameters for $K=4$, altering the original solution and allowing all elements (except Mg and Fe) to have contributions from additional processes. We find that S, Ca, C+N, Na, Cr, Mn, Co, Ni, and Ce are best fit with a third and/or fourth component, with such processes contributing most significantly at low metallicity. The $K=4$ model improves the ability of \name{} to fit the abundances of all elements but especially improves predictions of Ca, C+N, Fe, Mn, and Ce. This successful implementation of a $K=4$ model shows that \name{} can be extended to $K>2$, and it has potential future use in constraining enrichment beyond a single prompt and delayed process---critical to understanding enrichment from AGB stars, merging neutron stars, and rarer novae.

@@

\ejg{Some concluding statement about the tons of future science that KPM can do and how readers can access it.}

>>>>

% typesetting stuff
\shorttitle{a data-driven model for nucleosynthesis}
\shortauthors{Griffith \& Hogg}
\addtolength{\topmargin}{-0.4in} % trust the process
\addtolength{\textheight}{0.7in}
\setlength{\parindent}{1.6em}
\renewcommand{\twocolumngrid}{} % trust the process
\renewcommand{\paragraph}[1]{\bigskip\par\noindent{\textbf{#1}}~---}
\sloppy\sloppypar\raggedbottom\frenchspacing % trust the process

@@

\begin{abstract}\noindent % trust
Stellar surface abundances look like they are produced by two dominant processes, one prompt and one delayed.
We analyze the abundances of 14 elements in [X/Mg] vs. [Mg/H] space for 48,659 stars from APOGEE DR17 with a flexible, data-driven $K$-process model---dubbed \name.
In the fiducial model, where $K=2$, each elemental abundance of each star is described as the vector sum of a prompt (CCSN-like) and delayed (SNIa-like) process.
While prior work derives process vector components from median observed high-Ia and low-Ia abundance trends, we relax these assumptions and simultaneously fit all model parameters to the data, with only minimal constraints to keep the processes interpretable.
We find that our flexible method is able to recover the observed abundance bimodality and that this bimodality is visible to higher [Mg/H] in process amplitude space than in abundance space.
We compare fit parameters and statistics from this new method to those from prior work, finding that the two methods produce similar results, but that \name{} better predicts stellar abundances, especially for C+N and Mn.
We test the validity of the standard two-process model assumptions, such as that Fe is generated by a roughly half-half combination of prompt and delayed sources.
We find that the assumptions of prior works are generally confirmed, but that the inferred fractional contributions from each process are strongly dependent upon Fe assumptions.
We also explore third and fourth processes, calibrated to Ce and Mn, respectively. The additional processes improve the model's ability to predict the stellar abundances, demonstrating the model's ability to extend beyond $K=2$. \name{} enables future research to study the disk's two-dimensionallity, investigate age-abundance label correlations, fit non-bimodal abundance distributions with a multi-component nucleosynthetic model, and much more. \ejg{Add something about code availability.} 
\end{abstract}

@@

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data.
In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment.
The model is hierarchical, in that it learns some parameters (process vectors) that are shared across all stars, but different for each element, and some parameters (process amplitudes) that are shared across all elements, but different for each star.
The parameters output by our model can thus be used as de-noised abundance labels; these will sharpen relationships between abundances and stellar parameters (including birth location and time).

@@

\paragraph{1. $K$ processes}
All elements on the periodic table are produced by a combination of nucleosynthetic processes such as CCSN, SNIa, AGB stars, and merging neutron stars \citep{johnsonja2020}. The majority of $\alpha$, light odd-Z, and Fe-peak elements (the elements observed by APOGEE) are dominantly produced by $K=2$ sources, with one being a prompt process or mix of prompt processes, and one being a delayed process or a mix of delayed processes. This is substantiated by theoretical yields \citep[e.g.,][]{anderson2019, rybizki2017} and past successful data-driven models (e.g., G22, W22, \citealp{ting2022}). In this paper we therefore assume that $K \geq 2$, though \name{} could in principle be implemented with $K=1$.

\paragraph{2. Linearity}
At every metallicity, the (linear) $(X/H)$ abundances of a star can be expressed as a linear combination of $K$ processes.
These processes themselves will depend on metallicity, but a linear sum is sufficient to explain all element abundances at any overall metallicity.
Because yields depend on stellar masses and progenitor structure, and because different stars can get to their metallicities by different histories, this assumption must be at least slightly wrong in detail.

@@

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
The APOGEE errors may be slightly underestimated for some element measurements, so we add a softening parameter $Q$ to inflate the uncertainties on abundance-outlier stars (see Equation~\ref{eq:inflate_ivar}).

@@

While \name{} is flexible, we still include some quantitative assumptions, which we make to break exact model degeneracies (see Section~\ref{sec:model}).
Specifically, for each process, we choose one element to assign a ``known'' process vector element at all metallicities.
These are Mg and Fe in the 2-process case.
Specifically, for the prompt process, we ground our assumption in the nucleosynthetic theory that Mg is a pure CCSN element \citep[e.g.,][]{andrews2017}.
Unfortunately there is no comparable pure SNIa element, nor is there an element for which we know the relative CCSN/SNIa ratio.
In order to break an exact degeneracy between the prompt and delayed processes, we choose to fix the Fe process vector components, making unsubstantiated assumptions about Fe enrichment.

@@

We fit \name{} with $K=2$ to 15 abundances labels for 48,659 RGB stars in APOGEE DR17, selecting a population that minimizes statistical and systematic errors while spanning an [Mg/H] of -0.8 to 0.5. In the $K=2$ model, the first process, fixed to Mg, represents prompt CCSN-like enrichment, and the second process, fixed to Fe, represents delayed SNIa-like enrichment---but other nucleosynthetic sources with similar time delays may be mixed into each. Overall, we find that $K=2$ is a good fit to the data and that the model successfully recovers the global abundance patterns in the Milky Way. While \name{} does not rely on $\femg$ vs. $\mgh$ bimodality or median abundance trends, it is able to recover the observed bimodal abundance distribution. Further, the fit parameters $\Acc$ and $\AIa$ act as ``denoised'' abundance labels, and reveal a clearer signature of bimodality at high-metallicity in $\AIa/\Acc$ vs. $\mgh$ space than in $\femg$ vs. $\mgh$. This suggests that the \name{} fit parameters and predicted abundances could be used as a higher signal-to-noise tracer of nucleosynthesis, as they are condensing information from 15 elements into two variables.

@@

Finally, we expand \name{} from $K=2$ to $K=4$, regularizing the third and fourth processes to Ce and Mn. \name{} builds off of the original model, such that the $K=4$ model starts at the $K=2$ solution and then finds the best-fit parameters for $K=4$, altering the original solution and allowing all elements (except Mg and Fe) to have contributions from additional processes. We find that S, Ca, C+N, Na, Cr, Mn, Co, Ni, and Ce are best fit with a third and/or fourth component, with such processes contributing most significantly at low metallicity. The $K=4$ model improves the ability of \name{} to fit the abundances of all elements but especially improves predictions of Ca, C+N, Fe, Mn, and Ce. This successful implementation of a $K=4$ model shows that \name{} can be extended to $K>2$, and it has potential future use in constraining enrichment beyond a single prompt and delayed process---critical to understanding enrichment from AGB stars, merging neutron stars, and rarer novae.

@@

\ejg{Some concluding statement about the tons of future science that \name{} can do and how readers can access it.}
